
---
title: "Rolling Stone Dataset Analysis and Visualization"
---

# Rolling Stone Dataset Analysis

This document explores the Rolling Stone dataset using R. The analysis involves data cleaning, transformation, and visualization with `ggplot2`.

## Setting Up the Environment

We first load the necessary libraries and the dataset.

```{r setup}
# Required libraries
library(tidyverse)
library(gmodels)

# Load the dataset
data <- read.csv("rolling_stone.csv")
```

## Data Exploration and Transformation

### 1. Dataset Dimensions and Column Names

Retrieve the dataset's dimensions and column names.

```{r dataset-info}
data_info <- list(
  Dimensions = dim(data),
  Column_Names = colnames(data)
)
print("Dataset dimensions and column names:")
print(data_info)
```

### 2. Missing Data Analysis

Analyze the missing data in the dataset.

```{r missing-data}
missing_data <- data %>% summarise_all(~sum(is.na(.)))
print("Number of missing values:")
print(missing_data)

missing_data_percent <- data %>% summarise_all(~mean(is.na(.)) * 100)
print("Percentage of missing values (%):")
print(missing_data_percent)
```

### 3. Frequency Analysis of Categorical Variables

Summarize the frequency of categorical variables.

```{r categorical-summary}
categorical_summary <- data %>% 
  select(where(is.character)) %>% 
  summarise(across(everything(), ~list(table(.))))
print("Frequency analysis of categorical variables:")
print(categorical_summary)
```

### 4. Summary Statistics for Numerical Variables

Compute mean, median, and standard deviation for numerical variables.

```{r numerical-summary}
numerical_summary <- data %>% 
  select(where(is.numeric)) %>% 
  summarise_all(list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE)
  ))
print("Summary statistics for numerical variables:")
print(numerical_summary)
```

### 5. Top 5 Observations with the Lowest Rank in 2003

Identify the top 5 rows with the lowest rank in 2003.

```{r top-5-rows}
top_5_rows <- data %>% arrange(rank_2003) %>% head(5)
print("Top 5 observations with the lowest rank in 2003:")
print(top_5_rows)
```

### 6. Filter Data Based on Conditions

Filter rows based on specific conditions.

```{r filter-data}
filtered_data <- data %>% filter(rank_2012 > 50 & ave_age_at_top_500 < 100)
print("Data filtered based on specific conditions:")
print(filtered_data)
```

### 7. Add a New Column

Add a column calculating the rank difference between 2003 and 2020.

```{r add-column}
data <- data %>% mutate(rank_difference = rank_2003 - rank_2020)
print("New column (rank_difference) added:")
head(data)
```

### 8. Grouping and Averages

Group data by genre and calculate average Spotify popularity.

```{r grouping-averages}
grouped_data <- data %>% 
  group_by(genre) %>% 
  summarize(avg_popularity = mean(spotify_popularity, na.rm = TRUE))
print("Average Spotify popularity by genre:")
print(grouped_data)
```

### 9. Pivoting the Dataset

Transform the dataset from wide to long format.

```{r pivot-data}
long_data <- data %>% 
  pivot_longer(
    cols = c(rank_2003, rank_2012, rank_2020),
    names_to = "year",
    values_to = "rank"
  )
print("Pivoted dataset (long format):")
head(long_data)
```

### 10. Cross-Tabulation

Analyze the relationship between genre and artist gender.

```{r cross-tabulation}
cross_tab <- CrossTable(data$genre, data$artist_gender, prop.chisq = FALSE)
print("Relationship between categorical variables (cross-tabulation):")
print(cross_tab)
```

## Data Visualization

### 1. Bar Chart: Distribution of Genres

```{r bar-chart-genres}
data %>%
  count(genre) %>%
  ggplot(aes(x = reorder(genre, -n), y = n, fill = genre)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Distribution of Genres", x = "Genre", y = "Count") +
  theme_minimal()
```

### 2. Histogram: Distribution of Spotify Popularity

```{r histogram-popularity}
data %>%
  ggplot(aes(x = spotify_popularity)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Spotify Popularity", x = "Popularity", y = "Count") +
  theme_minimal()
```

### 3. Boxplot: Spotify Popularity by Genre

```{r boxplot-popularity-genre}
data %>%
  ggplot(aes(x = genre, y = spotify_popularity, fill = genre)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Spotify Popularity by Genre", x = "Genre", y = "Popularity") +
  theme_minimal()
```
# 4. Scatterplot: Rank 2003 vs. Rank 2020

```{r}
data %>%
  ggplot(aes(x = rank_2003, y = rank_2020)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Rank 2003 vs. Rank 2020", x = "Rank 2003", y = "Rank 2020") +
  theme_minimal()
```
# 5. Line plot: Average rank over years
```{r}
data_long <- data %>%
  pivot_longer(cols = c(rank_2003, rank_2012, rank_2020), names_to = "year", values_to = "rank")

data_long %>%
  group_by(year) %>%
  summarise(avg_rank = mean(rank, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = avg_rank, group = 1)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average Rank Over Years", x = "Year", y = "Average Rank") +
  theme_minimal()
```
# 6. Heatmap: Relationship between genre and artist gender

```{r}
data %>%
  count(genre, artist_gender) %>%
  ggplot(aes(x = genre, y = artist_gender, fill = n)) +
  geom_tile(color = "white") +
  labs(title = "Genre vs. Artist Gender", x = "Genre", y = "Artist Gender", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```